### 2025-01-21 paper/icml25/first/appendices/ablation.tex
refactoring

\section{Ablation Study}\label{sect:ablation}
While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.

@@

However, the performance gap narrows in certain scenarios. For instance, in the Dispersion scenario, \fnamer{} achieves comparable performance to \fname{}. This equivalence might be attributed to the increased complexity of learning the transition function in this particular environment.
\comment{GA: qui ho cambiato l'ablation perché ora mi sembra che shac++ in realtà vada meglio negli scenari aggiunti, che ne dici?}

>>>>

\section{Ablation Study}\label{apx:ablation}
While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.

@@

However, the performance gap narrows in certain scenarios. For instance, in the Dispersion scenario, \fnamer{} achieves comparable performance to \fname{}. This equivalence might be attributed to the increased complexity of learning the transition function in this particular environment.

### 2025-01-21 paper/icml25/first/appendices/scenarios.tex
refactoring

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
    \label{fig:experiments}
\end{figure*}

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
    \caption{Norm of policy gradients on the Transport environment across 5000 epochs. The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent. Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
\end{figure}

\section{Experiments}\label{sect:experiments}

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO. In particular, we choose the following scenario:
\begin{compactitem}
    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. 
        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward is $n$.
    \item \textbf{Transport} (\Cref{fig:transport}): 
        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. 
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
    \item \textbf{Discovery} (\Cref{fig:discovery}): 
        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. When $n=1$ we set $s=1$ and $k=7$. When $n>1$ we set $s=2$.
        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents. For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
     \item \textbf{Sampling} (\Cref{fig:sampling}): 
        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.  
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
\end{compactitem}

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}

\paragraph{Architectures}
We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture. Specifically, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 for the Policy Network and Value Network (for PPO, SHAC, and \fname{}). For the Reward Network, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 (for \fname{}). The Transition Network for \fname{} is a 3-layer single-head Transformer with hidden size 64.

While the MLP architecture is the simplest, the Transformer baseline benefits from position invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.

Overall, considering different training algorithms, architectures, scenarios, hyperparameters, we completed a total of $254$ runs lasting between $1$ to $8$ hours each.

\paragraph{Hardware Setup}
Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one Nvidia Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).

\subsection{Results}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

\paragraph{Dispersion}
In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable). In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation. In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior. In the third column (5 agents), PPO fails to converge, while \fname{} achieves high rewards (with no early stopping) but shows high variance due to increased environment complexity.

\paragraph{Discovery} 
The second row of \Cref{fig:experiments} shows a scenario with partial observability, namely discovery. In the first column (single agent), \fname{} outperforms PPO, though it does not reach the maximum reward, suggesting the world model struggles to capture dynamics. In the second and third columns (3 and 5 agents), PPO again fails to converge, whereas \fname{} maintains superior results, indicating cooperative behavior.

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes, with SHAC again more stable. 
With more agents, convergence is faster because the task becomes easier (more agents can push the package).
The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

\paragraph{Sampling}
The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC. In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC. In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes. Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.

>>>>

\section{Environments}\label{apx:scenarios}

\begin{compactitem}
    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. 
        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward is $n$.
    \item \textbf{Transport} (\Cref{fig:transport}): 
        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. 
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
    \item \textbf{Discovery} (\Cref{fig:discovery}): 
        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. When $n=1$ we set $s=1$ and $k=7$. When $n>1$ we set $s=2$.
        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents. For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
     \item \textbf{Sampling} (\Cref{fig:sampling}): 
        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.  
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
\end{compactitem}

### 2025-01-21 paper/icml25/first/sects/background.tex
refactoring

\newcommand{\agents}{\mathcal{N}}

@@

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: 
$$(\agents,\states,\{\observations{i}\}_{\scriptscriptstyle i\in\agents},\{\actions{i}\}_{\scriptscriptstyle i\in\agents},\{\stateobs{i}\}_{\scriptscriptstyle i\in\agents}, \{\rewards{i}\}_{\scriptscriptstyle i\in\agents}, \transition, \horizon, \discount, \initdist)$$
Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\observations{i}$ and $\actions{i}$ are the observation space and action space for the $i$-th agent ($i\in\agents$). $\stateobs{i}:\states\rightarrow\observations{i}$ is the projection to the observation space and $\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}$ is the reward function for the $i$-th agent. $\transition:\states\times\{\actions{i}\}_{i\in\agents}$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\{\actions{i}\}_{i\in\agents}\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent. Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space. We will denote the collective policy, determined by the collection of agent-specific policy simply with $\pi$. We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions. In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$. The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$. Finally, $s_0$ is sampled from the initial state distribution $\initdist$. For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.

@@

In this work, we will often approximate functions (such as transition $\transition$) with neural network. Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$. For instance, the transition and its corresponding neural network are denoted as $F$ and $F_{\theta}$, respectively. The corresponding loss will be denoted as $\mathcal{L}_\theta^g$ and it will aim to minimize the differences between $g$ and $g_\theta$.

@@

For simplicity, we formally treat only the case where the policy observe the full state to make the action, meaning that $\stateobs{i}$ is the identity function. However, in our experiments, we consider also scenarios where the full state is not available to the policies.

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

>>>>

@@

\subsection{Markov Decision Processes}
To provide context for our work, we first introduce the reinforcement learning (RL) framework. Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$. Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively. $\allo=\{\observations{i}\}_{i\in\agents}$ and $\alla=\{\actions{i}\}_{i\in\agents}$ are the observation space and action space for each agent. $\{\stateobs{i}:\states\rightarrow\observations{i}\}_{i\in\agents}$ are the projections from the state to the observation space and $\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}_{i\in\agents}$ are the reward functions for each agent. $\transition:\states\times\alla$ denotes the transition function. $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively. Finally, $\initdist$ is the initial state distribution.

Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent. Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space. We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi$. We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions. In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$. The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$. Finally, $s_0$ is sampled from the initial state distribution $\initdist$. For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.

@@

In this work, we will often approximate functions (such as transition $\transition$) with neural networks. Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$. For instance, the transition and its corresponding neural network are denoted as $F$ and $F_{\theta}$, respectively. The corresponding loss will be denoted as $\mathcal{L}_\theta^g$ and it will aim to minimize the differences between $g$ and $g_\theta$.

@@

For simplicity, we formally treat only the case where the policies observe the full state, meaning that $\stateobs{i}$ is the identity function. However, in our experiments, we consider also scenarios where the full state is not available to the policies.

@@

In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$. The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:  
\begin{equation*}\label{eq:shac}
    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
\end{equation*}
where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return. It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$. While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

### 2025-01-21 paper/icml25/first/sects/experiments.tex
minor on table 1
refactoring

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
    \label{fig:experiments}
\end{figure*}

@@

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO. In particular, we choose the following scenario:
\begin{compactitem}
    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. 
        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward is $n$.
    \item \textbf{Transport} (\Cref{fig:transport}): 
        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. 
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{World Model} The world model has enough information to predict the next states accurately.
        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
    \item \textbf{Discovery} (\Cref{fig:discovery}): 
        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. When $n=1$ we set $s=1$ and $k=7$. When $n>1$ we set $s=2$.
        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents. For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
     \item \textbf{Sampling} (\Cref{fig:sampling}): 
        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.  
        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
\end{compactitem}

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}\label{tab:max-rewards}
\end{table*}

\paragraph{Architectures}
We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture. Specifically, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 for the Policy Network and Value Network (for PPO, SHAC, and \fname{}). For the Reward Network, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 (for \fname{}). The Transition Network for \fname{} is a 3-layer single-head Transformer with hidden size 64.

While the MLP architecture is the simplest, the Transformer baseline benefits from position invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.

@@

\subsection{Results}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of normalized rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes, with SHAC again more stable. 
With more agents, convergence is faster because the task becomes easier (more agents can push the package).
The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

>>>>

\begin{figure*}[!t]
    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
    \label{fig:experiments}
\end{figure*}

@@

In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures. An ablation study when replacing only the transition function (meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO. In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling. For details regarding each scenario, see \Cref{apx:scenarios}. Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable. Transport and Sampling have differentiable reward functions, thus SHAC is applicable. For Transport and Dispersion, the observation space is complete wrt. the state. For Discovery and Sampling, the observation space is incomplete wrt. the state. In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC.

\begin{table*}[!t]
    \centering
    \input tables/max.tex
    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
\end{table*}

\paragraph{Architectures}
We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture. Specifically, we use a 1-layer MLP or a 1-layer single-head Transformer for policy, reward, and value network. The transition network used by \fname{} is always a 3-layer single-head Transformer

While the MLP architecture is the simplest, the Transformer baseline benefits from positional invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture. For more details on the architectures, see \Cref{apx:arch}.

@@

\subsection{Results}
In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

@@

\paragraph{Transport}
The third row of \Cref{fig:experiments} presents a differentiable scenario, transport. In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance. \fname{} converges faster, but SHAC is more consistent due to direct gradient stability. In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes. 
With more agents, convergence is faster because the task becomes easier (more agents can push the package).
The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package. This is a sign of generalization, as the agents learn to avoid collisions and push the package together.
